Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-12-05 response_to_reviewers.txt

We are especially grateful to Reviewer 2, who may have offered the most detailed and constructive review we have ever received.

Reviewer 1

I think that the abstract is too long and should be three-time shorter.
* The abstract is now substantially shorter. We have cut it down as much as we felt possible without losing clarity.

There are several occurrence of the word “taxon” or “taxa” while there is no definition: I guessed it was the leaves, and maybe the authors want to replace every occurrence.
* We replaced each occurrence of taxon or taxa with leaf or leaves.

### 2016-12-05 curvature.tex
added axis labels to access5 plots, revised text with R fix that was already in the response document

\begin{figure*}
	\includegraphics[width=\textwidth]{figs/access5-distribution.png}

@@

	\hspace*{\stretch{1}}
	\subfigure[\label{fig:5t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:6t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{\
        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-leaf trees with degree 44 and (b) 7-leaf trees with degree 70.
        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
				To limit clutter, these plots exclude pairs that are simple inclusions of pairs of trees with a smaller number of leaves (e.g. pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially.
        Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
				See online version for color figures; data tables available at \protect\url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
        }
	\label{fig:kappa-access}
\end{figure*}

@@

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
    %EM Don't you mean displaces $R$ from its original position?
	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $T$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Thus, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$.
	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

>>>>

\begin{figure*}
	\includegraphics[width=\textwidth]{figs/access5-distribution-alabels.png}

@@

	\hspace*{\stretch{1}}
	\subfigure[\label{fig:5t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:6t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{\
        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-leaf trees with degree 44 and (b) 7-leaf trees with degree 70.
        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
				To limit clutter, these plots exclude pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings) as their access time distributions do not change substantially.
        Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
				See online version for color figures; data tables available at \protect\url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
        }
	\label{fig:kappa-access}
\end{figure*}

@@

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
    %EM Don't you mean displaces $R$ from its original position?
	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $R$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Thus, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$.
	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}
